Stage-Splitting a Modal Language

### 2014-08-08 paper/splitting.tex
added more text to splitting

The input to our splitting algorithm is a term in \lang, and the output is two terms in \langmono, an
unstaged language.  The grammar for \langmono is given in \ref{fig:monoGrammar}.  It has no
staging features (we say that implicitly it has only one stage), and it is untyped.  
Modulo these differences, \langmono has all of the same features as \lang, 
although this is a matter of taste
\footnote{Since \lang is untyped, we could encode everything with just functions.}.

As with the dynamics, the form of the splitting statement depends on the external stage of the term.
We consider splitting for stage \bbtwo\ terms first, as it has a simpler form.  
For example, take \verb|hold{1+2} < 5|.
We can split this into two separate terms, \verb|1+2| and \verb|fn l => l < 5|, 
which are called the {\em precomputation} and {\em residual}, respectively.  
For this splitting to be correct, the residual applied to the precomputation should yield the same result as the original term.

@@

@@

We now discuss the formal definition of splitting.
The splitting algorithm comprises two judgments, $\splitonesym$ and $\splittwosym$.
The $\splitonesym$ judgment sends a stage \bbone\ term ($e$) to a combined term ($c$) and residual ($l.r$),
while $\splittwosym$ sends a stage \bbtwo\ term ($e$) to a precomputation ($p$) and a residual ($l.r$).
For concision, we actually represent residuals as a term open on a single variable, rather than as functions.
For both judgments, we use a context ($\Gamma$) to keep track of the open variables of $e$.

@@

We start with two careful definitions of equivalence ($\equiv$ on terms, $\sim$ on values).  They are mutually dependent.

@@

All I'm trying to say here is that two terms are equivalent if they evaluate to the same value,
where "same" for functions means that those functions always evaluate to the same thing.

>>>>

As with the dynamics, the form of the splitting statement depends on the external stage of the term.
We consider splitting for stage \bbtwo\ terms first, as it has a simpler form.  
For example, take \verb|hold{1+2} < 5|.
We can split this into two separate terms, \verb|1+2| and \verb|fn l => l < 5|, 
which are called the {\em precomputation} and {\em residual}
\footnote{We also used the term {\em residual} for the result of partial evaluation.
Where necessary, we will distinguish between these concepts by using the terms {\em splitting-residual}
and {\em evaluation-residual}.}, respectively.

@@

We must also develop a notion of what it means for splitting to be correct.  
Our general approach is to say that the dynamic semantics from \ref{sec:semantics} and splitting method are equivalent in some way.  
In general, if a multi-stage program $P$ splits into stage-\bbone\ part $P_1$ and stage-\bbtwo\ part $P_2$, then
\begin{enumerate}
\item Partial evaluation of $P$ will terminate if and only if evaluation of $P_1$ terminates, and if so then
\item the immediate results of each method will be equivalent, and
\item the evaluation-residual and the splitting-residual ($P_2$ bound under the precomputation) are equivalent.
\end{enumerate}

This is straight-forward, given a careful definition of equivalence.

@@

The splitting algorithm comprises two judgments, $\splitonesym$ and $\splittwosym$.
For both, the input is a term in \lang, and the output is two terms in \langmono, an
unstaged language.  The grammar for \langmono is given in \ref{fig:monoGrammar}.  It has no
staging features (we say that implicitly it has only one stage), and it is untyped.  
Modulo these differences, \langmono has all of the same features as \lang, 
although this is a matter of taste
\footnote{Since \lang is untyped, we could encode everything with just functions.}.

Specifically, the $\splitonesym$ judgment sends a stage \bbone\ term ($e$) to a combined term ($c$) and residual ($l.r$),
while $\splittwosym$ sends a stage \bbtwo\ term ($e$) to a precomputation ($p$) and a residual ($l.r$).
For concision, we represent residuals as a term open on a single variable, rather than as functions.
For both judgments, we use a context ($\Gamma$) to keep track of the open variables of $e$.

@@

We start with two mutually dependent definitions of equivalence.  
Both relate evaluation-residuals on the left with splitting residuals on the right,
but $\equiv$ operates on terms, whereas $\sim$ on values.

@@

Essentially, we can read these as saying that two terms are equivalent if they evaluate to the same value,
where "same" for functions means that those functions always evaluate to the same thing given equivalent inputs.
